---
title: "ST 558 Project 1"
author: "Chris Grace"
error: true
format: 
  html:
    toc: true
---

# Data Processing:
Before we can begin writing our helper function to query the census API, we need to understand the structure of its return. After obtaining an API key, we can use one of the provided example URLs:
```{r}
library(httr)

api_key <- "59c3be738e0d051c63b57b6327de0b11ef0f7a1e"

base_url <- paste0("https://api.census.gov/data/2022/acs/acs1/pums?get=SEX,PWGTP,MAR&for=state:*&SCHL=24&key=",
                   api_key)

api_return <- GET(base_url)

str(api_return)
```
The API call returns a lot of information, but it looks like the stuff we're interested in lives in the `content` element. We'll have to parse the JSON data before we can use it:

```{r}
suppressPackageStartupMessages(library(tidyverse))
library(jsonlite)

parsed <- fromJSON(rawToChar(api_return$content)) |>
          as_tibble(.name_repair = "minimal")

head(parsed)
```
After parsing the JSON data and formatting as a tibble, we can see that the first row returned are actually the variable names. We can fix this later.

## Helper Function
Now that we understand the type of object the API call returns, I think we're good to start building the helper function. Building our API call should be straightforward enough. Theoretically, we just use the `paste` function to build the URL, but we need to make sure each input the user provides is valid.

### Data Validation

Let's start by creating a barebones function with the given inputs, defaults, etc:
```{r}
query.census <- function(year = 2022, numeric_vars = "AGEP", categorical_vars = "SEX", 
                         geography = "State", specific_geo = "") {

}
```

We can validate the user's inputs one by one, stopping if the function is passed too few options, too many, or invalid ones

#### Validating Year

The user can input any year between 2010 and 2022 inclusive, but not 2020. Adding a check to our function:
```{r}
query.census <- function(year = 2022, numeric_vars = "AGEP", categorical_vars = "SEX", 
                         geography = "State", specific_geo = "") {
  # check for valid year
  if(year == 2020 | !(year %in% 2010:2022)) {
    stop("Error: Invalid year. Specify a year between 2010 and 2022 (not 2020)")
  } 
  
  # return message for testing
  return("API would be called")
}
```

Testing our validation:
```{r}
query.census()
query.census(2021)
query.census(2020)
query.census(-2000)
```

Working as intended!

#### Validating Numeric Variables

For numeric variables, we need to consider following:

-The user has to leave the default `AGEP` or specify their own selection of variables
-The additional variables they specify must be valid
-PWGTP should always be returned regardless of the user's selection
-There must be at least two numeric variables returned

To satisfy these conditions, we should:

1. Make sure the user doesn't pass `NULL` as their selection
2. Make sure each variable specified is valid
3. Add `PWGTP` to the vector of selected variables

Adding this logic to our function, we have:
```{r}
query.census <- function(year = 2022, numeric_vars = "AGEP", categorical_vars = "SEX", 
                         geography = "State", specific_geo = "") {
  # check for valid year
  if(year == 2020 | !(year %in% 2010:2022)) stop("Error: Invalid year. Specify a year between 2010 and 2022 (not 2020)")
  
  # check for valid numeric vars
  numeric_vars_opts <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  
  if(length(numeric_vars) == 0) stop("No numeric variable selected. Specify at least one.")
  
  for(i in numeric_vars) {
    if(!(i %in% numeric_vars_opts)) {
      stop("Invalid numeric variable. Specify a single variable or a vector of variables. Options: AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP")
    }
  }
  
  # make sure PWGTP ends up in final selection
  numeric_vars <- c(numeric_vars, "PWGTP")
  
  # return numeric vars for testing
  return(numeric_vars)
}
```

Let's test:
```{r}
query.census()
query.census(numeric_vars = NULL)
query.census(numeric_vars = "ST 558")
query.census(numeric_vars = c("JWDP", "GASP"))
```
Excellent!

#### Validating Categorical Variables

We can do the same thing for the categorical variables:
```{r}
query.census <- function(year = 2022, numeric_vars = "AGEP", categorical_vars = "SEX", 
                         geography = "State", specific_geo = "") {
  # check for valid year
  if(year == 2020 | !(year %in% 2010:2022)) stop("Error: Invalid year. Specify a year between 2010 and 2022 (not 2020)")
  
  # check for valid numeric vars
  numeric_vars_opts <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  if(length(numeric_vars) == 0) stop("No numeric variable selected. Specify at least one.")
  for(i in numeric_vars) {
    if(!(i %in% numeric_vars_opts)) stop("Invalid numeric variable. Specify a single variable or a vector of variables. Options: AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP")
  }
  
  numeric_vars <- c(numeric_vars, "PWGTP")
  
  # check for valid categorical vars
  categorical_vars_opts <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if(length(categorical_vars) == 0) stop("No categorical variable selected. Specify at least one or leave SEX as the default")
  for(i in categorical_vars) {
    if(!(i %in% categorical_vars_opts)) stop("Invalid categorical variable. Specify a single variable or a vector of variables. Options: FER, HHL, SCH, SCHL, SEX")
  }
  
  # return cat vars for testing
  return(categorical_vars)
}
```

More testing:
```{r}
query.census()
query.census(categorical_vars = c("SEX", "HHL"))
query.census(categorical_vars = "ST 558")
```

Perfect! The last thing to validate is the user's selection for geography. 

#### Validating Geography

We have to consider the following:

- `geography` must be **State**, **Region**, or **District**
- `specific_geo` must be between 1 and 50 for *state*, 1 and 4 for *region*, and 1 and 9 for *division*
- `specific_geo` can be "all" for any selection of `geography`

Adding these considerations to our function:
```{r}
query.census <- function(year = 2022, numeric_vars = "AGEP", categorical_vars = "SEX", 
                         geography = "State", specific_geo = 3) {
  # check for valid year
  if(year == 2020 | !(year %in% 2010:2022)) stop("Error: Invalid year. Specify a year between 2010 and 2022 (not 2020)")
  
  # check for valid numeric vars
  numeric_vars_opts <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  if(length(numeric_vars) == 0) stop("No numeric variable selected. Specify at least one.")
  for(i in numeric_vars) {
    if(!(i %in% numeric_vars_opts)) stop("Invalid numeric variable. Specify a single variable or a vector of variables. Options: AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP")
  }
  
  numeric_vars <- c(numeric_vars, "PWGTP")
  
  # check for valid categorical vars
  categorical_vars_opts <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if(length(categorical_vars) == 0) stop("No categorical variable selected. Specify at least one or leave SEX as the default")
  for(i in categorical_vars) {
    if(!(i %in% categorical_vars_opts)) stop("Invalid categorical variable. Specify a single variable or a vector of variables. Options: FER, HHL, SCH, SCHL, SEX")
  }
  
  # check for valid geography scope and region of interest
  geography_opts <- c("Region", "State", "Division")
  if(!(geography %in% geography_opts)) stop("Invalid geography classification. Specify one of: State, Region, Division")
  if(specific_geo == "all") {
    specific_geo = "*"
  }
  else {
    if(geography == "State" & !(specific_geo %in% 1:50)) stop("Invalid state requested. Specify a number 1 to 50 or `all`")
    if(geography == "Region" & !(specific_geo %in% 1:4)) stop("Invalid region requested. Specify a number 1 to 4 or `all`")
    if(geography == "Division" & !(specific_geo %in% 1:9)) stop("Invalid division requested. Specify a number 1 to 9 or `all`")
  }
  
  # return message for testing
  return("API would be called")
}
```

Our last round of data validation testing:
```{r}
query.census(geography = "State")
query.census(geography = "Region", specific_geo = 2)
query.census(geography = "ST 558", specific_geo = 2)
query.census(geography = "Division", specific_geo = 10)
query.census(geography = "Division", specific_geo = "all")
```
Looking good! Now that we know our input data is valid, we can build the URL and call the api from the function:

#### Building API Call
```{r}
query.census <- function(year = 2022, numeric_vars = "AGEP", categorical_vars = "SEX", 
                         geography = "State", specific_geo = 4) {
  suppressPackageStartupMessages(library(httr))
  suppressPackageStartupMessages(library(tidyverse))
  suppressPackageStartupMessages(library(jsonlite))
  
  # check for valid year
  if(year == 2020 | !(year %in% 2010:2022)) stop("Error: Invalid year. Specify a year between 2010 and 2022 (not 2020)")
  
  # check for valid numeric vars
  numeric_vars_opts <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  if(length(numeric_vars) == 0) stop("No numeric variable selected. Specify at least one.")
  for(i in numeric_vars) {
    if(!(i %in% numeric_vars_opts)) stop("Invalid numeric variable. Specify a single variable or a vector of variables. Options: AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP")
  }
  
  numeric_vars <- c(numeric_vars, "PWGTP")
  
  # check for valid categorical vars
  categorical_vars_opts <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if(length(categorical_vars) == 0) stop("No categorical variable selected. Specify at least one or leave SEX as the default")
  for(i in categorical_vars) {
    if(!(i %in% categorical_vars_opts)) stop("Invalid categorical variable. Specify a single variable or a vector of variables. Options: FER, HHL, SCH, SCHL, SEX")
  }
  
  # check for valid geography scope and region of interest
  geography_opts <- c("Region", "State", "Division")
  if(!(geography %in% geography_opts)) stop("Invalid geography classification. Specify one of: State, Region, Division")
  if(specific_geo == "all") {
    specific_geo = "*"
  }
  else {
    if(geography == "State" & !(specific_geo %in% 1:50)) stop("Invalid state requested. Specify a number 1 to 50 or `all`")
    if(geography == "Region" & !(specific_geo %in% 1:4)) stop("Invalid region requested. Specify a number 1 to 4 or `all`")
    if(geography == "Division" & !(specific_geo %in% 1:9)) stop("Invalid division requested. Specify a number 1 to 9 or `all`")
  }
  
  # build API URL
  api_key <- "59c3be738e0d051c63b57b6327de0b11ef0f7a1e"
  api_url <- paste0("https://api.census.gov/data/",
                    year,
                    "/acs/acs1/pums?get=",
                    paste0(numeric_vars, collapse = ","), # reformat numeric vars
                    ",",
                    paste0(categorical_vars, collapse = ","), # reformat categorical vars
                    "&for=",
                    tolower(geography), # region type
                    ":",
                    specific_geo, # specific region
                    "&key=",
                    api_key)
  
  api_return <- GET(api_url)
  api_parsed <- fromJSON(rawToChar(api_return$content)) |>
                as_tibble(.name_repair = "minimal")
  
  # return just a preview for testing
  return(head(api_parsed))
}
```

Let's test the default call (request SEX, AGEP, PWGTP for the default state (4 = Arizona))
```{r}
query.census()
```

While this is ugly data, ugly data is better than no data. This is the end of the data pipeline part, next we move onto cleaning.

### Data Cleaning

Next, we need to make sure the data is cleaned and coded correctly. As we can see, the first row of our data frame is the column names. Let's fix this:
```{r}
query.census <- function(year = 2022, numeric_vars = "AGEP", categorical_vars = "SEX", 
                         geography = "State", specific_geo = 4) {
  suppressPackageStartupMessages(library(httr))
  suppressPackageStartupMessages(library(tidyverse))
  suppressPackageStartupMessages(library(jsonlite))
  
  # check for valid year
  if(year == 2020 | !(year %in% 2010:2022)) stop("Error: Invalid year. Specify a year between 2010 and 2022 (not 2020)")
  
  # check for valid numeric vars
  numeric_vars_opts <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  if(length(numeric_vars) == 0) stop("No numeric variable selected. Specify at least one.")
  for(i in numeric_vars) {
    if(!(i %in% numeric_vars_opts)) stop("Invalid numeric variable. Specify a single variable or a vector of variables. Options: AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP")
  }
  
  numeric_vars <- c(numeric_vars, "PWGTP")
  
  # check for valid categorical vars
  categorical_vars_opts <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if(length(categorical_vars) == 0) stop("No categorical variable selected. Specify at least one or leave SEX as the default")
  for(i in categorical_vars) {
    if(!(i %in% categorical_vars_opts)) stop("Invalid categorical variable. Specify a single variable or a vector of variables. Options: FER, HHL, SCH, SCHL, SEX")
  }
  
  # check for valid geography scope and region of interest
  geography_opts <- c("Region", "State", "Division")
  if(!(geography %in% geography_opts)) stop("Invalid geography classification. Specify one of: State, Region, Division")
  if(specific_geo == "all") {
    specific_geo = "*"
  }
  else {
    if(geography == "State" & !(specific_geo %in% 1:50)) stop("Invalid state requested. Specify a number 1 to 50 or `all`")
    if(geography == "Region" & !(specific_geo %in% 1:4)) stop("Invalid region requested. Specify a number 1 to 4 or `all`")
    if(geography == "Division" & !(specific_geo %in% 1:9)) stop("Invalid division requested. Specify a number 1 to 9 or `all`")
  }
  
  # build API URL
  api_key <- "59c3be738e0d051c63b57b6327de0b11ef0f7a1e"
  api_url <- paste0("https://api.census.gov/data/",
                    year,
                    "/acs/acs1/pums?get=",
                    paste0(numeric_vars, collapse = ","), # reformat numeric vars
                    ",",
                    paste0(categorical_vars, collapse = ","), # reformat categorical vars
                    "&for=",
                    tolower(geography), # region type
                    ":",
                    specific_geo, # specific region
                    "&key=",
                    api_key)
  
  api_return <- GET(api_url)
  api_parsed <- fromJSON(rawToChar(api_return$content)) |>
                as_tibble()
  
  # fix col names
  output <- api_parsed
  names(output) <- output[1,]
  output <- output[-1,]
  
  # return just a preview to test func
  return(head(output))
}
```

Our columns have names now, but we still need to work on the *type* of each column. After looking at the dictionary, we should break up the variables into the following categories:

- **Factor**: SEX, FER, HHL, SCH, SCHL, State/Region/Division
- **Numeric**: PWGTP, AGEP, GASP, GRPIP, JWMNP
- **Time**: JWAP, JWDP

We'll worry about the time variables later. We can start by recasting the numeric and factor variables:
```{r}
query.census <- function(year = 2022, numeric_vars = "AGEP", categorical_vars = "SEX", 
                         geography = "State", specific_geo = 4) {
  suppressPackageStartupMessages(library(httr))
  suppressPackageStartupMessages(library(tidyverse))
  suppressPackageStartupMessages(library(jsonlite))
  
  # check for valid year
  if(year == 2020 | !(year %in% 2010:2022)) stop("Error: Invalid year. Specify a year between 2010 and 2022 (not 2020)")
  
  # check for valid numeric vars
  numeric_vars_opts <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  if(length(numeric_vars) == 0) stop("No numeric variable selected. Specify at least one.")
  for(i in numeric_vars) {
    if(!(i %in% numeric_vars_opts)) stop("Invalid numeric variable. Specify a single variable or a vector of variables. Options: AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP")
  }
  
  numeric_vars <- c(numeric_vars, "PWGTP")
  
  # check for valid categorical vars
  categorical_vars_opts <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if(length(categorical_vars) == 0) stop("No categorical variable selected. Specify at least one or leave SEX as the default")
  for(i in categorical_vars) {
    if(!(i %in% categorical_vars_opts)) stop("Invalid categorical variable. Specify a single variable or a vector of variables. Options: FER, HHL, SCH, SCHL, SEX")
  }
  
  # check for valid geography scope and region of interest
  geography_opts <- c("Region", "State", "Division")
  if(!(geography %in% geography_opts)) stop("Invalid geography classification. Specify one of: State, Region, Division")
  if(specific_geo == "all") {
    specific_geo = "*"
  }
  else {
    if(geography == "State" & !(specific_geo %in% 1:50)) stop("Invalid state requested. Specify a number 1 to 50 or `all`")
    if(geography == "Region" & !(specific_geo %in% 1:4)) stop("Invalid region requested. Specify a number 1 to 4 or `all`")
    if(geography == "Division" & !(specific_geo %in% 1:9)) stop("Invalid division requested. Specify a number 1 to 9 or `all`")
  }
  
  # build API URL
  api_key <- "59c3be738e0d051c63b57b6327de0b11ef0f7a1e"
  api_url <- paste0("https://api.census.gov/data/",
                    year,
                    "/acs/acs1/pums?get=",
                    paste0(numeric_vars, collapse = ","), # reformat numeric vars
                    ",",
                    paste0(categorical_vars, collapse = ","), # reformat categorical vars
                    "&for=",
                    tolower(geography), # region type
                    ":",
                    specific_geo, # specific region
                    "&key=",
                    api_key)
  
  api_return <- GET(api_url)
  api_parsed <- fromJSON(rawToChar(api_return$content)) |>
                as_tibble()
  
  # fix col names
  output <- api_parsed
  names(output) <- output[1,]
  output <- output[-1,]
  
  fctr_vars <- c("SEX", "FER", "HHL", "SCH", "SCHL", "state")
  numeric_vars <- c("PWGTP", 'AGEP', "GASP", "GRPIP", "JWMNP")
  time_Vars <- c("JWAP", "JWDP")
  
  # recast variables
  output <- output |> 
            mutate(across(any_of(numeric_vars), ~ as.numeric(.x)),
                   across(any_of(fctr_vars), ~ as.factor(.x)))

  # return just a preview to test func
  return(output)
}
```

Double checking to see if our recasting works:
```{r}
head(query.census())
```

The columns are the correct types, but we need to add labels to our factors.

### Assigning Factor Levels

Re-factoring each factor variable within our `query.census()` function will add a lot of clutter, so let's make another helper function to deal with all the factors.

While writing this function, I realized that the previous validation I did for the **State** is wrong. There are jumps and discontinuities in the state indices, so the list of acceptable values isn't just 1 to 50. In this factor helper function, I'm adding a parameter which will return the levels of **State** for use in the main helper function.
```{r}
level.fctrs <- function(input = NULL, return_state_levels = FALSE) {
  suppressPackageStartupMessages(library(dplyr))
  
  # level/labels for all factors
  SEX_levels <- c(1, 2)
  SEX_labels <- c("Male", "Female")
  FER_levels <- c(1, 2, 0)
  FER_labels <- c("Yes", "No", "Not Applicable")
  HHL_levels <- c(0, 1, 2, 3, 4, 5)
  HHL_labels <- c("Not Applicable", "English", "Spanish", "Other Indo-European Language",
                  "Asian and Pacific Island Language", "Other Language")
  SCH_levels <- c(0, 1, 2, 3)
  SCH_labels <- c("Not Applicable", "No", "Yes, public school/college", "Yes, private school/college")
  SCHL_levels <- c(0:24)
  SCHL_labels <- c("Not Applicable", "No Schooling", "Preschool", "Kindergarten", "Grade 1", "Grade 2",
                   "Grade 3", "Grade 4", "Grade 5", "Grade 6", "Grade 7", "Grade 8", "Grade 9", "Grade 10",
                   "Grade 11", "Grade 12 (No Diploma)", "High School Diploma", "GED/alternative diploma",
                   ">1 Year college", "1>= 1 Year college (No Degree)", "Associate's Degree", "Bachelor's Degree",
                   "Master's Degree", "Professional Degree (Beyond Bachelor's)", "Doctorate Degree")
  state_levels <- c(01, 02, 04, 05, 06, 08, 09, 11, 10, 12, 13, 15, 16, 17, 18, 19, 20, 21, 22, 23, 24, 25, 27, 
                    28, 26, 29, 30, 31, 32, 33, 34, 35, 36, 37, 38, 39, 40, 41, 42, 44, 45, 46, 47, 48, 49, 50, 
                    51, 53, 54, 55, 56)
  state_labels <- c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware",
                    "District of Columbia", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa",
                    "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota",
                    "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey",
                    "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon",
                    "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah",
                    "Vermont", "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming")
  region_levels <- c(1, 2, 3, 4)
  region_labels <- c("Northeast", "Mideast", "South", "West")
  division_levels <- c(1, 2, 3, 4, 5, 6, 7, 8, 9)
  division_labels <- c("New England", "Middle Atlantic", "East North Central", "West North Central", "South Atlantic", 
                       "East South Central", "West South Central", "Mountain", "Pacific")
  
  # just return state levels if requested
  if(return_state_levels) return(state_levels)
  
  # l33t metaprogramming 
  output <- input |>
              mutate(across(where(is.factor), ~ {
                var <- cur_column()
                levels <- get(paste0(var, "_levels"))
                labels <- get(paste0(var, "_labels"))
                factor(.x, levels = levels,
                           labels = labels)
              }))
  
  return(output)
}
```

This function expects an entire data frame/tibble, we can test it later.

### Fixing Time Variables

The time variables `JWAP` and `JWDP` are next on the chopping block. Looking at the docs, there are 286 values for each variable. Hard coding each level is a hard no. 

That being said, I do see a pattern - (almost) every time the variable(s) is incremented, the range is 5 minutes later. For some godforsaken reason, the range jumps 10 minutes when the variable goes from 7 -> 8 and from 10 -> 11. We can work around this quirk, but whoever is responsible deserves a lifetime of perpetually warm pillows. Shame.

Generally, if we set our clock to 12:02am (the middle range of the lowest value), we can add 5 times whatever value of `JWAP`/`JWDP` to get the time. I'm using `ifelse` as a sort of indicator variable if we think about this as a math function.

Let's make a helper function and test our hypothesis:
```{r}
fix.time <- function (time) {
  suppressPackageStartupMessages(library(lubridate))
  
  # don't mess up NA values
  is_zero <- time == 0
  
  scnds_since_midnight <- 120 + (time-1)*300 + ifelse(time > 6, 300, 0) + ifelse(time > 9, 300, 0)
  
  scnds_since_midnight[is_zero] <- 0
  
  # return vector converted to 24 hr time format
  return(hms::as_hms(scnds_since_midnight))
}
```

Now testing our hypothesis:

```{r}
fix.time(003) #12:12am (correct!)
fix.time(012) #01:07am (correct!)
fix.time(058) #05:02am (WRONG)
```
We have an issue. Checking again, there's another 10 minute interval when the variable goes from 20 -> 21. 
```{r}
fix.time <- function (time) {
  suppressPackageStartupMessages(library(lubridate))
  
  # don't mess up NA values
  is_zero <- time == 0
  
  scnds_since_midnight <- 120 + (time-1)*300 + ifelse(time > 6, 300, 0) + 
                          ifelse(time > 9, 300, 0) + ifelse(time > 20, 300, 0)
  
  scnds_since_midnight[is_zero] <- 0
  
  # return vector converted to 24 hr time format
  return(hms::as_hms(scnds_since_midnight))
}
```

Hopefully the last check:
```{r}
fix.time(058) # 5:02am (correct!)
fix.time(099) # 8:27am (correct!)
fix.time(234) # 7:42pm (correct!)
```

Alright! Now we can put the final touches on our helper function. I updated the following:

- Called `level.fctrs()` to assign labels to factor variables
- Called `fix.time()` to convert the weird time scheme to 24 hour format
- Updated the validation for the state indices

All together:
```{r}
query.census <- function(year = 2022, numeric_vars = "AGEP", categorical_vars = "SEX", 
                         geography = "State", specific_geo = 4) {
  suppressPackageStartupMessages(library(httr))
  suppressPackageStartupMessages(library(tidyverse))
  suppressPackageStartupMessages(library(jsonlite))
  
  # check for valid year
  if(year == 2020 | !(year %in% 2010:2022)) stop("Error: Invalid year. Specify a year between 2010 and 2022 (not 2020)")
  
  # check for valid numeric vars
  numeric_vars_opts <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  if(length(numeric_vars) == 0) stop("No numeric variable selected. Specify at least one.")
  for(i in numeric_vars) {
    if(!(i %in% numeric_vars_opts)) stop("Invalid numeric variable. Specify a single variable or a vector of variables. Options: AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP")
  }
  
  numeric_vars <- c(numeric_vars, "PWGTP")
  
  # check for valid categorical vars
  categorical_vars_opts <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if(length(categorical_vars) == 0) stop("No categorical variable selected. Specify at least one or leave SEX as the default")
  for(i in categorical_vars) {
    if(!(i %in% categorical_vars_opts)) stop("Invalid categorical variable. Specify a single variable or a vector of variables. Options: FER, HHL, SCH, SCHL, SEX")
  }
  
  # check for valid geography scope and region of interest
  geography_opts <- c("Region", "State", "Division")
  valid_states <- level.fctrs(return_state_levels = TRUE)
  if(!(geography %in% geography_opts)) stop("Invalid geography classification. Specify one of: State, Region, Division")
  if(specific_geo == "all") specific_geo = "*"
  else {
    if(geography == "State" & !(specific_geo %in% valid_states)) stop("Invalid state requested. Specify a valid state index or `all`")
    if(geography == "Region" & !(specific_geo %in% 1:4)) stop("Invalid region requested. Specify a number 1 to 4 or `all`")
    if(geography == "Division" & !(specific_geo %in% 1:9)) stop("Invalid division requested. Specify a number 1 to 9 or `all`")
  }
  
  # build API URL
  api_key <- "59c3be738e0d051c63b57b6327de0b11ef0f7a1e"
  api_url <- paste0("https://api.census.gov/data/",
                    year,
                    "/acs/acs1/pums?get=",
                    paste0(numeric_vars, collapse = ","), # reformat numeric vars
                    ",",
                    paste0(categorical_vars, collapse = ","), # reformat categorical vars
                    "&for=",
                    tolower(geography), # region type
                    ":",
                    specific_geo, # specific region
                    "&key=",
                    api_key)
  
  api_return <- GET(api_url)
  api_parsed <- fromJSON(rawToChar(api_return$content)) |>
                as_tibble()
  
  # fix col names
  output <- api_parsed
  names(output) <- output[1,]
  output <- output[-1,]
  
  fctr_vars <- c("SEX", "FER", "HHL", "SCH", "SCHL", "state", "region", "division")
  numeric_vars <- c("PWGTP", 'AGEP', "GASP", "GRPIP", "JWMNP")
  time_vars <- c("JWAP", "JWDP")
  
  # recast variables
  output <- output |> 
            mutate(across(any_of(c("region", "state", "division")), ~ as.numeric(.x)),
                   across(any_of(numeric_vars), ~ as.numeric(.x)),
                   across(any_of(time_vars), ~ as.numeric(.x) |> fix.time()),
                   across(any_of(fctr_vars), ~ as.factor(.x))) |>
            level.fctrs()

  return(output)
}
```

Let's test our function! We'll start by looking at the 2021 data for North Carolina (state 37):
```{r}
nc_output <- query.census(year = 2021, specific_geo = 37)
head(nc_output)
```

Looks great! Let's try adding some more variables and selecting a division (9 = Pacific)
```{r}
pac_output <- query.census(year = 2021, numeric_vars = c("AGEP", "GASP", "JWAP"),
                           categorical_vars = c("SEX", "HHL"), geography = "Division",
                           specific_geo = 9)

head(pac_output)
```

It looks like there's an issue with the `JWAP` and `HHL` columns. I'm adding a new argument to the `query.census` function to return the URL (instead of the table). This makes it easier to check for discrepencies if something seems off.

### "Final" Helper Function

```{r}
query.census <- function(year = 2022, numeric_vars = "AGEP", categorical_vars = "SEX", 
                         geography = "State", specific_geo = 4, return_url = FALSE) {
  suppressPackageStartupMessages(library(httr))
  suppressPackageStartupMessages(library(tidyverse))
  suppressPackageStartupMessages(library(jsonlite))
  
  # check for valid year
  if(year == 2020 | !(year %in% 2010:2022)) stop("Error: Invalid year. Specify a year between 2010 and 2022 (not 2020)")
  
  # check for valid numeric vars
  numeric_vars_opts <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  if(length(numeric_vars) == 0) stop("No numeric variable selected. Specify at least one.")
  for(i in numeric_vars) {
    if(!(i %in% numeric_vars_opts)) stop("Invalid numeric variable. Specify a single variable or a vector of variables. Options: AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP")
  }
  
  numeric_vars <- c(numeric_vars, "PWGTP")
  
  # check for valid categorical vars
  categorical_vars_opts <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if(length(categorical_vars) == 0) stop("No categorical variable selected. Specify at least one or leave SEX as the default")
  for(i in categorical_vars) {
    if(!(i %in% categorical_vars_opts)) stop("Invalid categorical variable. Specify a single variable or a vector of variables. Options: FER, HHL, SCH, SCHL, SEX")
  }
  
  # check for valid geography scope and region of interest
  geography_opts <- c("Region", "State", "Division")
  valid_states <- level.fctrs(return_state_levels = TRUE)
  if(!(geography %in% geography_opts)) stop("Invalid geography classification. Specify one of: State, Region, Division")
  if(specific_geo == "all") specific_geo = "*"
  else {
    if(geography == "State" & !(specific_geo %in% valid_states)) stop("Invalid state requested. Specify a valid state index or `all`")
    if(geography == "Region" & !(specific_geo %in% 1:4)) stop("Invalid region requested. Specify a number 1 to 4 or `all`")
    if(geography == "Division" & !(specific_geo %in% 1:9)) stop("Invalid division requested. Specify a number 1 to 9 or `all`")
  }
  
  # build API URL
  api_key <- "59c3be738e0d051c63b57b6327de0b11ef0f7a1e"
  api_url <- paste0("https://api.census.gov/data/",
                    year,
                    "/acs/acs1/pums?get=",
                    paste0(numeric_vars, collapse = ","), # reformat numeric vars
                    ",",
                    paste0(categorical_vars, collapse = ","), # reformat categorical vars
                    "&for=",
                    tolower(geography), # region type
                    ":",
                    specific_geo, # specific region
                    "&key=",
                    api_key)
  
  if(return_url) return(api_url)
  
  api_return <- GET(api_url)
  api_parsed <- fromJSON(rawToChar(api_return$content)) |> 
                as_tibble(.name_repair = "minimal")
  
  # fix col names
  output <- api_parsed
  names(output) <- output[1,]
  output <- output[-1,]
  
  fctr_vars <- c("SEX", "FER", "HHL", "SCH", "SCHL", "state", "region", "division")
  numeric_vars <- c("PWGTP", 'AGEP', "GASP", "GRPIP", "JWMNP")
  time_vars <- c("JWAP", "JWDP")
  
  # recast variables
  output <- output |> 
            mutate(across(any_of(c("region", "state", "division")), ~ as.numeric(.x)),
                   across(any_of(numeric_vars), ~ as.numeric(.x)),
                   across(any_of(time_vars), ~ as.numeric(.x) |> fix.time()),
                   across(any_of(fctr_vars), ~ as.factor(.x))) |>
            level.fctrs()
  
  return(output)
}
```

After going to the URL of the query, it seems like everything is working correctly on our end. The raw values for these observations are indeed 0/NA for most of the `JWAP` and `HHL` values in this sample. This means that we're done with the helper function!

## Calling Multiple Years

To get data for multiple years, we can simply call the `query.census()` function once for each year and combine the results. Within the `for()` loop we add the `year` variable.
```{r}
query.census.multi <- function(years = 2010:2022, numeric_vars = "AGEP", categorical_vars = "SEX", 
                               geography = "State", specific_geo = 4, return_url = FALSE) {
  if(min(years) < 2010 | max(years) > 2022) {
    stop("Error: Invalid year(s). Specify year(s) between 2010 and 2022")
  } else if(2020 %in% years) {
    warning("Warning: 2020 included in year selection. 2020 Data is not inluded in the output.")
  }
  
  output <- data.frame()
  
  # never include 2020 regardless of user selection
  for(i in years[years != 2020]) {
    year_data <- query.census(year = i, numeric_vars = numeric_vars, categorical_vars = categorical_vars,
                              geography = geography, specific_geo = specific_geo)
    
    year_data$year <- i
    
    output <- rbind(output, year_data)
  }
  
  return(output)  
}
```

Testing:
```{r}
test_multi <- query.census.multi(2019:2021)
head(test_multi)
unique(test_multi$year)
```
No 2020! Perfect.

# Summary Function
Before we make the summary function, let's create a new class for the object returned by our helper function:
```{r}
query.census <- function(year = 2022, numeric_vars = "AGEP", categorical_vars = "SEX", 
                         geography = "State", specific_geo = 4, return_url = FALSE) {
  suppressPackageStartupMessages(library(httr))
  suppressPackageStartupMessages(library(tidyverse))
  suppressPackageStartupMessages(library(jsonlite))
  
  # check for valid year
  if(year == 2020 | !(year %in% 2010:2022)) stop("Error: Invalid year. Specify a year between 2010 and 2022 (not 2020)")
  
  # check for valid numeric vars
  numeric_vars_opts <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  if(length(numeric_vars) == 0) stop("No numeric variable selected. Specify at least one.")
  for(i in numeric_vars) {
    if(!(i %in% numeric_vars_opts)) stop("Invalid numeric variable. Specify a single variable or a vector of variables. Options: AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP")
  }
  
  numeric_vars <- c(numeric_vars, "PWGTP")
  
  # check for valid categorical vars
  categorical_vars_opts <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if(length(categorical_vars) == 0) stop("No categorical variable selected. Specify at least one or leave SEX as the default")
  for(i in categorical_vars) {
    if(!(i %in% categorical_vars_opts)) stop("Invalid categorical variable. Specify a single variable or a vector of variables. Options: FER, HHL, SCH, SCHL, SEX")
  }
  
  # check for valid geography scope and region of interest
  geography_opts <- c("Region", "State", "Division")
  valid_states <- level.fctrs(return_state_levels = TRUE)
  if(!(geography %in% geography_opts)) stop("Invalid geography classification. Specify one of: State, Region, Division")
  if(specific_geo == "all") specific_geo = "*"
  else {
    if(geography == "State" & !(specific_geo %in% valid_states)) stop("Invalid state requested. Specify a valid state index or `all`")
    if(geography == "Region" & !(specific_geo %in% 1:4)) stop("Invalid region requested. Specify a number 1 to 4 or `all`")
    if(geography == "Division" & !(specific_geo %in% 1:9)) stop("Invalid division requested. Specify a number 1 to 9 or `all`")
  }
  
  # build API URL
  api_key <- "59c3be738e0d051c63b57b6327de0b11ef0f7a1e"
  api_url <- paste0("https://api.census.gov/data/",
                    year,
                    "/acs/acs1/pums?get=",
                    paste0(numeric_vars, collapse = ","), # reformat numeric vars
                    ",",
                    paste0(categorical_vars, collapse = ","), # reformat categorical vars
                    "&for=",
                    tolower(geography), # region type
                    ":",
                    specific_geo, # specific region
                    "&key=",
                    api_key)
  
  if(return_url) return(api_url)
  
  api_return <- GET(api_url)
  api_parsed <- fromJSON(rawToChar(api_return$content)) |> 
                as_tibble(.name_repair = "minimal")
  
  # fix col names
  output <- api_parsed
  names(output) <- output[1,]
  output <- output[-1,]
  
  fctr_vars <- c("SEX", "FER", "HHL", "SCH", "SCHL", "state", "region", "division")
  numeric_vars <- c("PWGTP", 'AGEP', "GASP", "GRPIP", "JWMNP")
  time_vars <- c("JWAP", "JWDP")
  
  # recast variables
  output <- output |> 
            mutate(across(any_of(c("region", "state", "division")), ~ as.numeric(.x)),
                   across(any_of(numeric_vars), ~ as.numeric(.x)),
                   across(any_of(time_vars), ~ as.numeric(.x) |> fix.time()),
                   across(any_of(fctr_vars), ~ as.factor(.x))) |>
            level.fctrs()
  
  # reclassify output
  class(output) <- c("census", class(output))
  
  return(output)
}
```

Checking to make sure the function returns an object of class `census`:
```{r}
test_census <- query.census()

class(test_census)
```

Great! Now we can work on our summary function.

The first step is validating the user's variable selections. We need to ensure:

- The default selection is all numeric and categorical variables
- Every variable the user specifies is in the data they provide
- We do not include `year`, `JWAP`, and `JWDP` in our summaries
- We do not include `PWGTP` in our summary (but we use it to compute them)

I'm using the `setdiff()` function to check that the user's variable selections are present in the input data frame. If the function has a non-zero length result, we know there's an invalid selection.
```{r}
summary.census <- function(census, numeric_vars = "all", categorical_vars = "all") {
  valid_numeric <- names(select(census, where(is.numeric)))
  valid_categorical <- names(select(census, where(is.factor)))
  
  # validate numeric vars
  if(numeric_vars == "all") {
    numeric_vars = valid_numeric
  } else if(length(setdiff(numeric_vars, valid_numeric)) > 0) {
    stop("At least one numeric variable requested is not in the census data provided")
  }
  
  # validate categorical vars
  if(categorical_vars == "all") {
    categorical_vars = valid_categorical
  } else if(length(setdiff(categorical_vars, valid_categorical)) > 0) {
    stop("At least one categorical variable requested is not in the census data provided")
  }
  
  # test return
  return(c("function would run, numeric vars: ", numeric_vars))
}
```

Testing our function:
```{r}
summary(test_census)
summary(test_census, numeric_vars = "ST 558")
summary(test_census, numeric_vars = "AGEP", categorical_vars = "SEX")
```
Great! Next, we need to exclude any of the variables we aren't summarizing, even if the user requests them.

Again, the `setdiff()` operation comes in handy:

```{r}
summary.census <- function(census, numeric_vars = "all", categorical_vars = "all") {
  valid_numeric <- names(select(census, where(is.numeric)))
  valid_categorical <- names(select(census, where(is.factor)))
  invalid_vars <- c("PWGTP", "JWAP", "JWDP", "year")
  
  # validate numeric vars
  if(numeric_vars == "all") {
    numeric_vars = valid_numeric
  } else if(length(setdiff(numeric_vars, valid_numeric)) > 0) {
    stop("At least one numeric variable requested is not in the census data provided")
  }
  
  # remove PWGTP/year from numeric vars to summarize
  numeric_vars <- setdiff(numeric_vars, invalid_vars)
    
  # validate categorical vars
  if(categorical_vars == "all") {
    categorical_vars = valid_categorical
  } else if(length(setdiff(categorical_vars, valid_categorical)) > 0) {
    stop("At least one categorical variable requested is not in the census data provided")
  }
  
  # test return
  return(c("function would run, numeric vars: ", numeric_vars))
}
```

Testing to make sure `PWGTP` isn't included:
```{r}
summary(test_census)
```

Excellent! Now, we just need to compute and return the numeric/categorical variable stats:
```{r}
summary.census <- function(census, numeric_vars = "all", categorical_vars = "all") {
  valid_numeric <- names(select(census, where(is.numeric)))
  valid_categorical <- names(select(census, where(is.factor)))
  invalid_vars <- c("PWGTP", "JWAP", "JWDP", "year")
  
  # validate numeric vars
  if(numeric_vars == "all") {
    numeric_vars = valid_numeric
  } else if(length(setdiff(numeric_vars, valid_numeric)) > 0) {
    stop("At least one numeric variable requested is not in the census data provided")
  }
  
  # remove PWGTP/year from numeric vars to summarizr
  numeric_vars <- setdiff(numeric_vars, invalid_vars)
    
  # validate categorical vars
  if(categorical_vars == "all") {
    categorical_vars = valid_categorical
  } else if(length(setdiff(categorical_vars, valid_categorical)) > 0) {
    stop("At least one categorical variable requested is not in the census data provided")
  }
  
  # compute numeric stats
  numeric_stats <- census |>
                   summarize(across(all_of(numeric_vars), list(
                          mean = ~ sum(.x*PWGTP)/sum(PWGTP),
                          sd = ~ sqrt(sum((.x-(sum(.x*PWGTP)/sum(PWGTP)))^2 * PWGTP) / sum(PWGTP))),
                          .names = "{.col}_{.fn}"))
  
  # doo doo horrible convoluted way of computing categorical stats (it works though)
  categorical_stats <- census |>
                       select(all_of(categorical_vars)) |>
                       pivot_longer(everything(), names_to = "var", values_to = "level") |>
                       count(var, level, name = "n") |>
                       mutate(colname = paste0(var, "_", level, "_n")) |>
                       select(colname, n) |>
                       pivot_wider(names_from = colname, values_from = n)
  
  # return results
  return(c(numeric_stats, categorical_stats))
}
```

The way I computed the counts at each level of the categorical variables is incredibly clunky. More importantly though, it works. Let's test our function:
```{r}
summary(test_census)
```
Nice! But we should try a larger table:
```{r}
big_test <- query.census.multi(years = 2010:2011, numeric_vars = c("AGEP", "JWMNP", "JWAP"),
                               categorical_vars = c("HHL", "SEX"), geography = "Region",
                               specific_geo = "all")

# small scale test first
summary(big_test, numeric_vars = "JWMNP", categorical_vars = "SEX")

# large scale test
summary(big_test)
```
Wait, it worked? Most of this project has been fistfights with my computer but this function actually worked first try. Beautiful.

## Fixing Time Helper Function

One of my group mates asked if the coding for `JWAP` was different than for `JWDP`. I realized I never checked for a difference because clearly they'd be the same, right? WRONG. My existing function works for `JWAP` but we'll need to make a new one for `JWDP`. I'm going to add an argument to my `fix.time()` function to specify which is being computed. Looking at the values `JWDP` takes on, it seems more borked than `JWAP`. The pattern seems to be:

- 1-6: 30 min jumps
- 7-18: 10 min jumps
- 19-78: 5 min jumps
- 79-132: 10 min jumps
- 133-136: 30 min jumps
- 137-148: 10 min jumps
- 148-150: 30 min jumps

What possesses someone to do this? I have zero idea of why you'd choose to record data like this. Anyways:
```{r}
fix.time <- function (time, type) {
  library(lubridate)
  
  is_zero <- time == 0
  
  if(type == "JWAP") {
    scnds_since_midnight <- 120 + (time-1)*300 + ifelse(time > 6, 150, 0) + 
                            ifelse(time > 7, 150, 0) + ifelse(time > 9, 150, 0) +
                            ifelse(time > 10, 150, 0) + ifelse(time > 20, 150, 0) +
                            ifelse(time > 21, 150, 0)
  } else if(type == "JWDP") {
    scnds_since_midnight <- 900 + min(5, time-1)*1800 + 
           ifelse(time > 6, 1170 + min(11, time - 7)*600, 0) +
           ifelse(time > 18, 450 + min(59, time - 19)*300, 0) +
           ifelse(time > 78, 450 + min(53, time - 79)*600, 0) +
           ifelse(time > 132, 1200 + min(3, time - 133)*1800, 0) +
           ifelse(time > 136, 1200 + min(11, time - 137)*600, 0) +
           ifelse(time > 148, 1200 + min(2, time - 149)*1800, 0)
  } else {
    stop("Invalid type entered. Enter either 'JWAP' or 'JWDP'.")
  }
  
  scnds_since_midnight[is_zero] <- 0
  
  return(hms::as_hms(scnds_since_midnight))
}
```

I also made an adjustment to how I calculate the `JWAP` times so that the midpoint is still correct even for the irregularly large ranges.

## Fixing the Helper Function
We need to make a small change to our helper function now that `fix.time()` requires two arguments:
```{r}
query.census <- function(year = 2022, numeric_vars = "AGEP", categorical_vars = "SEX", 
                         geography = "State", specific_geo = 4, return_url = FALSE) {
  suppressPackageStartupMessages(library(httr))
  suppressPackageStartupMessages(library(tidyverse))
  suppressPackageStartupMessages(library(jsonlite))
  
  # check for valid year
  if(year == 2020 | !(year %in% 2010:2022)) stop("Error: Invalid year. Specify a year between 2010 and 2022 (not 2020)")
  
  # check for valid numeric vars
  numeric_vars_opts <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  if(length(numeric_vars) == 0) stop("No numeric variable selected. Specify at least one.")
  for(i in numeric_vars) {
    if(!(i %in% numeric_vars_opts)) stop("Invalid numeric variable. Specify a single variable or a vector of variables. Options: AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP")
  }
  
  numeric_vars <- c(numeric_vars, "PWGTP")
  
  # check for valid categorical vars
  categorical_vars_opts <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if(length(categorical_vars) == 0) stop("No categorical variable selected. Specify at least one or leave SEX as the default")
  for(i in categorical_vars) {
    if(!(i %in% categorical_vars_opts)) stop("Invalid categorical variable. Specify a single variable or a vector of variables. Options: FER, HHL, SCH, SCHL, SEX")
  }
  
  # check for valid geography scope and region of interest
  geography_opts <- c("Region", "State", "Division")
  valid_states <- level.fctrs(return_state_levels = TRUE)
  if(!(geography %in% geography_opts)) stop("Invalid geography classification. Specify one of: State, Region, Division")
  if(specific_geo == "all") specific_geo = "*"
  else {
    if(geography == "State" & !(specific_geo %in% valid_states)) stop("Invalid state requested. Specify a valid state index or `all`")
    if(geography == "Region" & !(specific_geo %in% 1:4)) stop("Invalid region requested. Specify a number 1 to 4 or `all`")
    if(geography == "Division" & !(specific_geo %in% 1:9)) stop("Invalid division requested. Specify a number 1 to 9 or `all`")
  }
  
  # build API URL
  api_key <- "59c3be738e0d051c63b57b6327de0b11ef0f7a1e"
  api_url <- paste0("https://api.census.gov/data/",
                    year,
                    "/acs/acs1/pums?get=",
                    paste0(numeric_vars, collapse = ","), # reformat numeric vars
                    ",",
                    paste0(categorical_vars, collapse = ","), # reformat categorical vars
                    "&for=",
                    tolower(geography), # region type
                    ":",
                    specific_geo, # specific region
                    "&key=",
                    api_key)
  
  if(return_url) return(api_url)
  
  api_return <- GET(api_url)
  api_parsed <- fromJSON(rawToChar(api_return$content)) |> 
                as_tibble(.name_repair = "minimal")
  
  # fix col names
  output <- api_parsed
  names(output) <- output[1,]
  output <- output[-1,]
  
  fctr_vars <- c("SEX", "FER", "HHL", "SCH", "SCHL", "state", "region", "division")
  numeric_vars <- c("PWGTP", 'AGEP', "GASP", "GRPIP", "JWMNP")
  time_vars <- c("JWAP", "JWDP")
  
  # recast variables
  output <- output |> 
            mutate(across(any_of(c("region", "state", "division")), ~ as.numeric(.x)),
                   across(any_of(numeric_vars), ~ as.numeric(.x)),
                   across(any_of(fctr_vars), ~ as.factor(.x)),
                   across(any_of("JWAP"), ~ as.numeric(.x) |> fix.time(type = "JWAP")),
                   across(any_of("JWDP"), ~ as.numeric(.x) |> fix.time(type = "JWDP"))) |>
            level.fctrs()
  
  # reclassify output
  class(output) <- c("census", class(output))
  
  return(output)
}
```

# Plotting Function
Lastly, we'll make a quick function to produce box plots. As usual, we start with data validation to ensure:

- The user doesn't specify variables not present in the data
- The user doesn't specify a non-plotting variable (`JWAP`/`JWDP`, `PWGTP`)
- The user doesn't specify multiple variables

Most of this logic can just be carried down from our summary function:
```{r}
plot.census <- function(census, numeric_var, categorical_var) {
  suppressPackageStartupMessages(library(dplyr))
  
  valid_numeric <- names(select(census, where(is.numeric)))
  valid_categorical <- names(select(census, where(is.factor)))
  invalid_vars <- c("PWGTP", "JWAP", "JWDP")
  
  # validate numeric vars
  if(length(numeric_var) > 1) {
    stop("Multiple numeric variables specified. Specify only one.")
  } else if(!(numeric_var %in% valid_numeric)) {
    stop("The numeric variable requested is not in the census data provided")
  } else if(numeric_var %in% invalid_vars) {
    stop("The numeric variable requested is invalid. Invalid variables: JWAP, JWDP, PWGTP")
  }
    
  # validate categorical vars
  if(length(categorical_var) > 1) {
    stop("Multiple categorical variables specified. Specify only one.")
  } else if(!(categorical_var %in% valid_categorical)) {
    stop("At least one categorical variable requested is not in the census data provided")
  }
  
  return(numeric_var)
}
```

A quick test:

```{r}
plot(test_census, "JWAP", "SEX")
plot(test_census, c("ST 558", "AGEP"), "SEX")
plot(test_census, "AGEP", "SEX")
```
Great! Now we just write a quick function to plot the two variables:
```{r}
plot.census <- function(census, numeric_var, categorical_var) {
  suppressPackageStartupMessages(library(dplyr))
  library(ggplot2)
  
  valid_numeric <- names(select(census, where(is.numeric)))
  valid_categorical <- names(select(census, where(is.factor)))
  invalid_vars <- c("PWGTP", "JWAP", "JWDP")
  
  # validate numeric vars
  if(length(numeric_var) > 1) {
    stop("Multiple numeric variables specified. Specify only one.")
  } else if(!(numeric_var %in% valid_numeric)) {
    stop("The numeric variable requested is not in the census data provided")
  } else if(numeric_var %in% invalid_vars) {
    stop("The numeric variable requested is invalid. Invalid variables: JWAP, JWDP, PWGTP")
  }
    
  # validate categorical vars
  if(length(categorical_var) > 1) {
    stop("Multiple categorical variables specified. Specify only one.")
  } else if(!(categorical_var %in% valid_categorical)) {
    stop("At least one categorical variable requested is not in the census data provided")
  }
  
  # build title string
  title <- paste("Distribution of",
                 numeric_var,
                 "by",
                 categorical_var)
  
  # plot!
  plot <- ggplot(census, aes(y = get(numeric_var), 
                             fill = get(categorical_var), 
                             weight = PWGTP)) +
          geom_boxplot() +
          labs(title = title,
               x = "",
               y = numeric_var,
               fill = categorical_var) +
          scale_x_discrete(labels = NULL, breaks = NULL) +
          theme_minimal()
  
  return(plot)
}
```

Testing:
```{r}
plot(test_census, "AGEP", "SEX")

# subsetted data!!!
plot(big_test[1:100000,], "JWMNP", categorical_var = "region")
```
Beautiful! Note that the above boxplot only reflects the first 100000 observations.

# Investigating Education and Work Commute

Suppose we have a research question:

**Do American's with more education spend more time commuting to work?**

The Census would be a reliable source of data to answer this question. As a first step, let's check out the most recent data for North Carolina (2022):
```{r}
nc_2022 <- query.census(numeric_vars = "JWMNP", categorical_vars = "SCH",
                        geography = "State", specific_geo = 37)

summary(nc_2022)
```
There are many zero values of `JWMNP` in our dataset (reflecting `N/A` values). We can subset our data to include observations where `JWMNP` is nonzero:
```{r}
nc_2022_sub <- nc_2022 |>
               filter(JWMNP > 0)

summary(nc_2022_sub)
```
These values look more accurate, as the average work commute nationally is between 20 and 30 minutes.

The summary function shows us the counts for each level of education and the mean/sd for work commute time, but we're actually concerned with the relationship between them. 

Let's create a box plot with our `plot()` function:
```{r}
plot(nc_2022_sub, numeric_var = "JWMNP", categorical_var = "SCH")
```

The box plot shows an identical median time of 20 minutes and Q3 time of 32 minutes for all education levels. There are many upper outliers for the "no" education group, with fewer for the "Yes, public school" group and even fewer for the "Yes, private school" group. The second quartile for the "No" group is narrower than the other education levels.

The next step is to acquire data for the whole country. Let's acquire data from 2021 and 2022 this time:
```{r}
usa <- query.census.multi(2021:2022, categorical_vars = "SCH", numeric_vars = "JWMNP",
                          geography = "Region", specific_geo = "all")
summary(usa)
```
Our mean commute time is very similar to the NC data. Let's remove the values of zero and re-run the summary:
```{r}
usa_sub <- usa |>
           filter(JWMNP > 0)

summary(usa_sub)
```
It looks like the national mean commute time is also similar to the NC data. I tried to plot the data, it loaded for half an hour before I decided we needed an even smaller subset.

The code below samples `n` observations from our census data to speed up plotting. I'm sampling with replacement so we're quasi-bootstrapping.
```{r}
# runtime: ~ 1 year
#plot(usa_sub, "JWMNP", "SCH")

n <- 100000
samples <- sample(1:nrow(usa_sub), size = n, replace = TRUE)

usa_sampled <- usa_sub[samples,]

plot(usa_sampled, "JWMNP", "SCH")
```
The distribution of commute times nationally is very similar to the distribution in North Carolina. We lack the evidence to answer a definitive yes/no to our research question if we only look at the box plot and numerical summaries. A regression model may be useful here, but at least it seems like our functions work properly!

# Final Functions:
## Factor-Fixer:
```{r}
level.fctrs <- function(input = NULL, return_state_levels = FALSE) {
  suppressPackageStartupMessages(library(dplyr))
  
  # level/labels for all factors
  SEX_levels <- c(1, 2)
  SEX_labels <- c("Male", "Female")
  FER_levels <- c(1, 2, 0)
  FER_labels <- c("Yes", "No", "Not Applicable")
  HHL_levels <- c(0, 1, 2, 3, 4, 5)
  HHL_labels <- c("Not Applicable", "English", "Spanish", "Other Indo-European Language",
                  "Asian and Pacific Island Language", "Other Language")
  SCH_levels <- c(0, 1, 2, 3)
  SCH_labels <- c("Not Applicable", "No", "Yes, public school/college", "Yes, private school/college")
  SCHL_levels <- c(0:24)
  SCHL_labels <- c("Not Applicable", "No Schooling", "Preschool", "Kindergarten", "Grade 1", "Grade 2",
                   "Grade 3", "Grade 4", "Grade 5", "Grade 6", "Grade 7", "Grade 8", "Grade 9", "Grade 10",
                   "Grade 11", "Grade 12 (No Diploma)", "High School Diploma", "GED/alternative diploma",
                   ">1 Year college", "1>= 1 Year college (No Degree)", "Associate's Degree", "Bachelor's Degree",
                   "Master's Degree", "Professional Degree (Beyond Bachelor's)", "Doctorate Degree")
  state_levels <- c(01, 02, 04, 05, 06, 08, 09, 11, 10, 12, 13, 15, 16, 17, 18, 19, 20, 21, 22, 23, 24, 25, 27, 
                    28, 26, 29, 30, 31, 32, 33, 34, 35, 36, 37, 38, 39, 40, 41, 42, 44, 45, 46, 47, 48, 49, 50, 
                    51, 53, 54, 55, 56)
  state_labels <- c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware",
                    "District of Columbia", "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa",
                    "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota",
                    "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey",
                    "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon",
                    "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah",
                    "Vermont", "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming")
  region_levels <- c(1, 2, 3, 4)
  region_labels <- c("Northeast", "Mideast", "South", "West")
  division_levels <- c(1, 2, 3, 4, 5, 6, 7, 8, 9)
  division_labels <- c("New England", "Middle Atlantic", "East North Central", "West North Central", "South Atlantic", 
                       "East South Central", "West South Central", "Mountain", "Pacific")
  
  # just return state levels if requested
  if(return_state_levels) return(state_levels)
  
  # l33t metaprogramming 
  output <- input |>
              mutate(across(where(is.factor), ~ {
                var <- cur_column()
                levels <- get(paste0(var, "_levels"))
                labels <- get(paste0(var, "_labels"))
                factor(.x, levels = levels,
                           labels = labels)
              }))
  
  return(output)
}
```


## Helper Function
```{r}
query.census <- function(year = 2022, numeric_vars = "AGEP", categorical_vars = "SEX", 
                         geography = "State", specific_geo = 4, return_url = FALSE) {
  suppressPackageStartupMessages(library(httr))
  suppressPackageStartupMessages(library(tidyverse))
  suppressPackageStartupMessages(library(jsonlite))
  
  # check for valid year
  if(year == 2020 | !(year %in% 2010:2022)) stop("Error: Invalid year. Specify a year between 2010 and 2022 (not 2020)")
  
  # check for valid numeric vars
  numeric_vars_opts <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  if(length(numeric_vars) == 0) stop("No numeric variable selected. Specify at least one.")
  for(i in numeric_vars) {
    if(!(i %in% numeric_vars_opts)) stop("Invalid numeric variable. Specify a single variable or a vector of variables. Options: AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP")
  }
  
  numeric_vars <- c(numeric_vars, "PWGTP")
  
  # check for valid categorical vars
  categorical_vars_opts <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if(length(categorical_vars) == 0) stop("No categorical variable selected. Specify at least one or leave SEX as the default")
  for(i in categorical_vars) {
    if(!(i %in% categorical_vars_opts)) stop("Invalid categorical variable. Specify a single variable or a vector of variables. Options: FER, HHL, SCH, SCHL, SEX")
  }
  
  # check for valid geography scope and region of interest
  geography_opts <- c("Region", "State", "Division")
  valid_states <- level.fctrs(return_state_levels = TRUE)
  if(!(geography %in% geography_opts)) stop("Invalid geography classification. Specify one of: State, Region, Division")
  if(specific_geo == "all") specific_geo = "*"
  else {
    if(geography == "State" & !(specific_geo %in% valid_states)) stop("Invalid state requested. Specify a valid state index or `all`")
    if(geography == "Region" & !(specific_geo %in% 1:4)) stop("Invalid region requested. Specify a number 1 to 4 or `all`")
    if(geography == "Division" & !(specific_geo %in% 1:9)) stop("Invalid division requested. Specify a number 1 to 9 or `all`")
  }
  
  # build API URL
  api_key <- "59c3be738e0d051c63b57b6327de0b11ef0f7a1e"
  api_url <- paste0("https://api.census.gov/data/",
                    year,
                    "/acs/acs1/pums?get=",
                    paste0(numeric_vars, collapse = ","), # reformat numeric vars
                    ",",
                    paste0(categorical_vars, collapse = ","), # reformat categorical vars
                    "&for=",
                    tolower(geography), # region type
                    ":",
                    specific_geo, # specific region
                    "&key=",
                    api_key)
  
  if(return_url) return(api_url)
  
  api_return <- GET(api_url)
  api_parsed <- fromJSON(rawToChar(api_return$content)) |> 
                as_tibble(.name_repair = "minimal")
  
  # fix col names
  output <- api_parsed
  names(output) <- output[1,]
  output <- output[-1,]
  
  fctr_vars <- c("SEX", "FER", "HHL", "SCH", "SCHL", "state", "region", "division")
  numeric_vars <- c("PWGTP", 'AGEP', "GASP", "GRPIP", "JWMNP")
  time_vars <- c("JWAP", "JWDP")
  
  # recast variables
  output <- output |> 
            mutate(across(any_of(c("region", "state", "division")), ~ as.numeric(.x)),
                   across(any_of(numeric_vars), ~ as.numeric(.x)),
                   across(any_of(fctr_vars), ~ as.factor(.x)),
                   across(any_of("JWAP"), ~ as.numeric(.x) |> fix.time(type = "JWAP")),
                   across(any_of("JWDP"), ~ as.numeric(.x) |> fix.time(type = "JWDP"))) |>
            level.fctrs()
  
  # reclassify output
  class(output) <- c("census", class(output))
  
  return(output)
}
```

## Multi-Year Function
```{r}
query.census.multi <- function(years = 2010:2022, numeric_vars = "AGEP", categorical_vars = "SEX", 
                               geography = "State", specific_geo = 4, return_url = FALSE) {
  if(min(years) < 2010 | max(years) > 2022) {
    stop("Error: Invalid year(s). Specify year(s) between 2010 and 2022")
  } else if(2020 %in% years) {
    warning("Warning: 2020 included in year selection. 2020 Data is not inluded in the output.")
  }
  
  output <- data.frame()
  
  # never include 2020 regardless of user selection
  for(i in years[years != 2020]) {
    year_data <- query.census(year = i, numeric_vars = numeric_vars, categorical_vars = categorical_vars,
                              geography = geography, specific_geo = specific_geo)
    
    year_data$year <- i
    
    output <- rbind(output, year_data)
  }
  
  return(output)  
}
```

## Time Fixing Helper Function
```{r}
fix.time <- function (time, type) {
  library(lubridate)
  
  is_zero <- time == 0
  
  if(type == "JWAP") {
    scnds_since_midnight <- 120 + (time-1)*300 + ifelse(time > 6, 150, 0) + 
                            ifelse(time > 7, 150, 0) + ifelse(time > 9, 150, 0) +
                            ifelse(time > 10, 150, 0) + ifelse(time > 20, 150, 0) +
                            ifelse(time > 21, 150, 0)
  } else if(type == "JWDP") {
    scnds_since_midnight <- 900 + min(5, time-1)*1800 + 
           ifelse(time > 6, 1170 + min(11, time - 7)*600, 0) +
           ifelse(time > 18, 450 + min(59, time - 19)*300, 0) +
           ifelse(time > 78, 450 + min(53, time - 79)*600, 0) +
           ifelse(time > 132, 1200 + min(3, time - 133)*1800, 0) +
           ifelse(time > 136, 1200 + min(11, time - 137)*600, 0) +
           ifelse(time > 148, 1200 + min(2, time - 149)*1800, 0)
  } else {
    stop("Invalid type entered. Enter either 'JWAP' or 'JWDP'.")
  }
  
  scnds_since_midnight[is_zero] <- 0
  
  return(hms::as_hms(scnds_since_midnight))
}
```

## Summary Function
```{r}
summary.census <- function(census, numeric_vars = "all", categorical_vars = "all") {
  valid_numeric <- names(select(census, where(is.numeric)))
  valid_categorical <- names(select(census, where(is.factor)))
  invalid_vars <- c("PWGTP", "JWAP", "JWDP", "year")
  
  # validate numeric vars
  if(numeric_vars == "all") {
    numeric_vars = valid_numeric
  } else if(length(setdiff(numeric_vars, valid_numeric)) > 0) {
    stop("At least one numeric variable requested is not in the census data provided")
  }
  
  # remove PWGTP/year from numeric vars to summarizr
  numeric_vars <- setdiff(numeric_vars, invalid_vars)
    
  # validate categorical vars
  if(categorical_vars == "all") {
    categorical_vars = valid_categorical
  } else if(length(setdiff(categorical_vars, valid_categorical)) > 0) {
    stop("At least one categorical variable requested is not in the census data provided")
  }
  
  # compute numeric stats
  numeric_stats <- census |>
                   summarize(across(all_of(numeric_vars), list(
                          mean = ~ sum(.x*PWGTP)/sum(PWGTP),
                          sd = ~ sqrt(sum((.x-(sum(.x*PWGTP)/sum(PWGTP)))^2 * PWGTP) / sum(PWGTP))),
                          .names = "{.col}_{.fn}"))
  
  # doo doo horrible convoluted way of computing categorical stats (it works though)
  categorical_stats <- census |>
                       select(all_of(categorical_vars)) |>
                       pivot_longer(everything(), names_to = "var", values_to = "level") |>
                       count(var, level, name = "n") |>
                       mutate(colname = paste0(var, "_", level, "_n")) |>
                       select(colname, n) |>
                       pivot_wider(names_from = colname, values_from = n)
  
  # return results
  return(c(numeric_stats, categorical_stats))
}
```

## Plotting Function
```{r}
plot.census <- function(census, numeric_var, categorical_var) {
  suppressPackageStartupMessages(library(dplyr))
  library(ggplot2)
  
  valid_numeric <- names(select(census, where(is.numeric)))
  valid_categorical <- names(select(census, where(is.factor)))
  invalid_vars <- c("PWGTP", "JWAP", "JWDP")
  
  # validate numeric vars
  if(length(numeric_var) > 1) {
    stop("Multiple numeric variables specified. Specify only one.")
  } else if(!(numeric_var %in% valid_numeric)) {
    stop("The numeric variable requested is not in the census data provided")
  } else if(numeric_var %in% invalid_vars) {
    stop("The numeric variable requested is invalid. Invalid variables: JWAP, JWDP, PWGTP")
  }
    
  # validate categorical vars
  if(length(categorical_var) > 1) {
    stop("Multiple categorical variables specified. Specify only one.")
  } else if(!(categorical_var %in% valid_categorical)) {
    stop("At least one categorical variable requested is not in the census data provided")
  }
  
  # build title string
  title <- paste("Distribution of",
                 numeric_var,
                 "by",
                 categorical_var)
  
  # plot!
  plot <- ggplot(census, aes(y = get(numeric_var), 
                             fill = get(categorical_var), 
                             weight = PWGTP)) +
          geom_boxplot() +
          labs(title = title,
               x = "",
               y = numeric_var,
               fill = categorical_var) +
          scale_x_discrete(labels = NULL, breaks = NULL) +
          theme_minimal()
  
  return(plot)
}
```

