---
title: "ST 558 Project 1"
author: "Chris Grace"
format: html
error: true
---

# Data Processing:
Before we can begin writing our helper function to query the census API, we need to understand the structure of its return. After obtaining an API key, we can use one of the provided example URLs:
```{r}
library(httr)

api_key <- "59c3be738e0d051c63b57b6327de0b11ef0f7a1e"

base_url <- paste0("https://api.census.gov/data/2022/acs/acs1/pums?get=SEX,PWGTP,MAR&for=state:*&SCHL=24&key=",
                   api_key)

api_return <- GET(base_url)

str(api_return)
```
The API call returns a lot of information, but it looks like the stuff we're interested in lives in the `content` element. We'll have to parse the JSON data before we can use it:

```{r}
suppressPackageStartupMessages(library(tidyverse))
library(jsonlite)

parsed <- fromJSON(rawToChar(api_return$content)) |>
          as_tibble()

head(parsed)
```
After parsing the JSON data and formatting as a tibble, we can see that the first row returned are actually the variable names. We can fix this later.

## Helper Function
Now that we understand the type of object the API call returns, I think we're good to start building the helper function. Building our API call should be straightforward enough (theoretically we just use the `paste` function), but we need to make sure each input the user provides is valid.

### Data Validation

Let's start by creating a barebones function with the given inputs, defaults, etc:
```{r}
query.census <- function(year = 2022, numeric_vars = "AGEP", categorical_vars = "SEX", 
                         geography = "State", specific_geo = "") {

}
```

We can validate the user's inputs one by one, stopping if the function is passed too few options, too many, or invalid ones. We can start by validating the user entered a correct year. They can input any year between 2010 and 2022 inclusive, but not 2020. Adding a check to our function:
```{r}
query.census <- function(year = 2022, numeric_vars = "AGEP", categorical_vars = "SEX", 
                         geography = "State", specific_geo = "") {
  # check for valid year
  if(year == 2020 | !(year %in% 2010:2022)) stop("Error: Invalid year. Specify a year between 2010 and 2022 (not 2020)")
  
  return("API would be called")
}
```

Let's make sure the year validation is working as intended. For testing, I just have the function return a message if the input is valid.
```{r}
query.census()
query.census(2021)
query.census(2020)
query.census(-2000)
```

Working as intended! Next is validating the inputs for the numeric and categorical variables. 

For numeric variables, we need to consider following:

-The user has to leave the default `AGEP` or specify their own selection of variables
-The additional variables they specify must be valid
-PWGTP should always be returned regardless of the user's selection
-There must be at least two numeric variables returned

To satisfy these conditions, we should:

1. Make sure the user doesn't pass `NULL` as their selection
2. Make sure each variable specified is valid
3. Add `PWGTP` to the vector of selected variables

Adding this workflow to our function, we have the following. I'm returning the vector of numeric variables to verify everything is working correctly:
```{r}
query.census <- function(year = 2022, numeric_vars = "AGEP", categorical_vars = "SEX", 
                         geography = "State", specific_geo = "") {
  # check for valid year
  if(year == 2020 | !(year %in% 2010:2022)) stop("Error: Invalid year. Specify a year between 2010 and 2022 (not 2020)")
  
  # check for valid numeric vars
  numeric_vars_opts <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  if(length(numeric_vars) == 0) stop("No numeric variable selected. Specify at least one.")
  for(i in numeric_vars) {
    if(!(i %in% numeric_vars_opts)) stop("Invalid numeric variable. Specify a single variable or a vector of variables. Options: AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP")
  }
  
  numeric_vars <- c(numeric_vars, "PWGTP")
  
  return(numeric_vars)
}
```

Let's test:

```{r}
query.census()
query.census(numeric_vars = NULL)
query.census(numeric_vars = "ST 558")
query.census(numeric_vars = c("JWDP", "GASP"))
```
Working as intended! We can do the same thing for the categorical variables:
```{r}
query.census <- function(year = 2022, numeric_vars = "AGEP", categorical_vars = "SEX", 
                         geography = "State", specific_geo = "") {
  # check for valid year
  if(year == 2020 | !(year %in% 2010:2022)) stop("Error: Invalid year. Specify a year between 2010 and 2022 (not 2020)")
  
  # check for valid numeric vars
  numeric_vars_opts <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  if(length(numeric_vars) == 0) stop("No numeric variable selected. Specify at least one.")
  for(i in numeric_vars) {
    if(!(i %in% numeric_vars_opts)) stop("Invalid numeric variable. Specify a single variable or a vector of variables. Options: AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP")
  }
  
  numeric_vars <- c(numeric_vars, "PWGTP")
  
  # check for valid categorical vars
  categorical_vars_opts <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if(length(categorical_vars) == 0) stop("No categorical variable selected. Specify at least one or leave SEX as the default")
  for(i in categorical_vars) {
    if(!(i %in% categorical_vars_opts)) stop("Invalid categorical variable. Specify a single variable or a vector of variables. Options: FER, HHL, SCH, SCHL, SEX")
  }
  
  return(categorical_vars)
}
```

More testing:
```{r}
query.census()
query.census(categorical_vars = c("SEX", "HHL"))
query.census(categorical_vars = "ST 558")
```

Perfect! The last thing to validate is the user's selection for geography. We have to consider the following:

- `geography` must be **State**, **Region**, or **District**
- `specific_geo` must be between 1 and 50 for state, 1 and 4 for region, and 1 and 9 for division
- `specific_geo` can be "all" for any selection of `geography`

Adding these considerations to our function:
```{r}
query.census <- function(year = 2022, numeric_vars = "AGEP", categorical_vars = "SEX", 
                         geography = "State", specific_geo = 3) {
  # check for valid year
  if(year == 2020 | !(year %in% 2010:2022)) stop("Error: Invalid year. Specify a year between 2010 and 2022 (not 2020)")
  
  # check for valid numeric vars
  numeric_vars_opts <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  if(length(numeric_vars) == 0) stop("No numeric variable selected. Specify at least one.")
  for(i in numeric_vars) {
    if(!(i %in% numeric_vars_opts)) stop("Invalid numeric variable. Specify a single variable or a vector of variables. Options: AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP")
  }
  
  numeric_vars <- c(numeric_vars, "PWGTP")
  
  # check for valid categorical vars
  categorical_vars_opts <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if(length(categorical_vars) == 0) stop("No categorical variable selected. Specify at least one or leave SEX as the default")
  for(i in categorical_vars) {
    if(!(i %in% categorical_vars_opts)) stop("Invalid categorical variable. Specify a single variable or a vector of variables. Options: FER, HHL, SCH, SCHL, SEX")
  }
  
  # check for valid geography scope and region of interest
  geography_opts <- c("Region", "State", "Division")
  if(!(geography %in% geography_opts)) stop("Invalid geography classification. Specify one of: State, Region, Division")
  if(specific_geo == "all") {
    specific_geo = "*"
  }
  else {
    if(geography == "State" & !(specific_geo %in% 1:50)) stop("Invalid state requested. Specify a number 1 to 50 or `all`")
    if(geography == "Region" & !(specific_geo %in% 1:4)) stop("Invalid region requested. Specify a number 1 to 4 or `all`")
    if(geography == "Division" & !(specific_geo %in% 1:9)) stop("Invalid division requested. Specify a number 1 to 9 or `all`")
  }
  
  return("API would be called")
}
```

Our last round of data validation testing:
```{r}
query.census(geography = "State")
query.census(geography = "Region", specific_geo = 2)
query.census(geography = "ST 558", specific_geo = 2)
query.census(geography = "Division", specific_geo = 10)
query.census(geography = "Division", specific_geo = "all")
```
Looking good! Now that we know our input data is valid, we can build the URL and call the api from the function:

```{r}
query.census <- function(year = 2022, numeric_vars = "AGEP", categorical_vars = "SEX", 
                         geography = "State", specific_geo = 4) {
  # check for valid year
  if(year == 2020 | !(year %in% 2010:2022)) stop("Error: Invalid year. Specify a year between 2010 and 2022 (not 2020)")
  
  # check for valid numeric vars
  numeric_vars_opts <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  if(length(numeric_vars) == 0) stop("No numeric variable selected. Specify at least one.")
  for(i in numeric_vars) {
    if(!(i %in% numeric_vars_opts)) stop("Invalid numeric variable. Specify a single variable or a vector of variables. Options: AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP")
  }
  
  numeric_vars <- c(numeric_vars, "PWGTP")
  
  # check for valid categorical vars
  categorical_vars_opts <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if(length(categorical_vars) == 0) stop("No categorical variable selected. Specify at least one or leave SEX as the default")
  for(i in categorical_vars) {
    if(!(i %in% categorical_vars_opts)) stop("Invalid categorical variable. Specify a single variable or a vector of variables. Options: FER, HHL, SCH, SCHL, SEX")
  }
  
  # check for valid geography scope and region of interest
  geography_opts <- c("Region", "State", "Division")
  if(!(geography %in% geography_opts)) stop("Invalid geography classification. Specify one of: State, Region, Division")
  if(specific_geo == "all") {
    specific_geo = "*"
  }
  else {
    if(geography == "State" & !(specific_geo %in% 1:50)) stop("Invalid state requested. Specify a number 1 to 50 or `all`")
    if(geography == "Region" & !(specific_geo %in% 1:4)) stop("Invalid region requested. Specify a number 1 to 4 or `all`")
    if(geography == "Division" & !(specific_geo %in% 1:9)) stop("Invalid division requested. Specify a number 1 to 9 or `all`")
  }
  
  api_key <- "59c3be738e0d051c63b57b6327de0b11ef0f7a1e"
  api_url <- paste0("https://api.census.gov/data/",
                    year,
                    "/acs/acs1/pums?get=",
                    paste0(numeric_vars, collapse = ","), # reformat numeric vars
                    ",",
                    paste0(categorical_vars, collapse = ","), # reformat categorical vars
                    "&for=",
                    tolower(geography), # region type
                    ":",
                    specific_geo, # specific region
                    "&key=",
                    api_key)
  
  api_return <- GET(api_url)
  api_parsed <- fromJSON(rawToChar(api_return$content)) |>
                as_tibble()
  
  # return just a preview to test func
  return(head(api_parsed))
}
```

Let's test the function by calling it's default (request SEX, AGEP, PWGTP for the default state (4 = Arizona))
```{r}
query.census()
```

Great! Next, we need to make sure the data is cleaned and coded correctly. As we can see, the first row is the column names. Let's fix this:
```{r}
query.census <- function(year = 2022, numeric_vars = "AGEP", categorical_vars = "SEX", 
                         geography = "State", specific_geo = 4) {
  # check for valid year
  if(year == 2020 | !(year %in% 2010:2022)) stop("Error: Invalid year. Specify a year between 2010 and 2022 (not 2020)")
  
  # check for valid numeric vars
  numeric_vars_opts <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  if(length(numeric_vars) == 0) stop("No numeric variable selected. Specify at least one.")
  for(i in numeric_vars) {
    if(!(i %in% numeric_vars_opts)) stop("Invalid numeric variable. Specify a single variable or a vector of variables. Options: AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP")
  }
  
  numeric_vars <- c(numeric_vars, "PWGTP")
  
  # check for valid categorical vars
  categorical_vars_opts <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if(length(categorical_vars) == 0) stop("No categorical variable selected. Specify at least one or leave SEX as the default")
  for(i in categorical_vars) {
    if(!(i %in% categorical_vars_opts)) stop("Invalid categorical variable. Specify a single variable or a vector of variables. Options: FER, HHL, SCH, SCHL, SEX")
  }
  
  # check for valid geography scope and region of interest
  geography_opts <- c("Region", "State", "Division")
  if(!(geography %in% geography_opts)) stop("Invalid geography classification. Specify one of: State, Region, Division")
  if(specific_geo == "all") {
    specific_geo = "*"
  }
  else {
    if(geography == "State" & !(specific_geo %in% 1:50)) stop("Invalid state requested. Specify a number 1 to 50 or `all`")
    if(geography == "Region" & !(specific_geo %in% 1:4)) stop("Invalid region requested. Specify a number 1 to 4 or `all`")
    if(geography == "Division" & !(specific_geo %in% 1:9)) stop("Invalid division requested. Specify a number 1 to 9 or `all`")
  }
  
  api_key <- "59c3be738e0d051c63b57b6327de0b11ef0f7a1e"
  api_url <- paste0("https://api.census.gov/data/",
                    year,
                    "/acs/acs1/pums?get=",
                    paste0(numeric_vars, collapse = ","), # reformat numeric vars
                    ",",
                    paste0(categorical_vars, collapse = ","), # reformat categorical vars
                    "&for=",
                    tolower(geography), # region type
                    ":",
                    specific_geo, # specific region
                    "&key=",
                    api_key)
  
  api_return <- GET(api_url)
  api_parsed <- fromJSON(rawToChar(api_return$content)) |>
                as_tibble()
  
  # fix col names
  output <- api_parsed
  names(output) <- output[1,]
  output <- output[-1,]
  
  # return just a preview to test func
  return(head(output))
}
```

Our columns have names now, but we still need to work on the types of each column. After looking at the dictionary, we should break up the variables into the following categories:

- Factor: SEX, FER, HHL, SCH, SCHL
- Numeric: PWGTP, AGEP, GASP, GRPIP, JWMNP
- Time: JWAP, JWDP

```{r}
query.census <- function(year = 2022, numeric_vars = "AGEP", categorical_vars = "SEX", 
                         geography = "State", specific_geo = 4) {
  # check for valid year
  if(year == 2020 | !(year %in% 2010:2022)) stop("Error: Invalid year. Specify a year between 2010 and 2022 (not 2020)")
  
  # check for valid numeric vars
  numeric_vars_opts <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP")
  if(length(numeric_vars) == 0) stop("No numeric variable selected. Specify at least one.")
  for(i in numeric_vars) {
    if(!(i %in% numeric_vars_opts)) stop("Invalid numeric variable. Specify a single variable or a vector of variables. Options: AGEP, GASP, GRPIP, JWAP, JWDP, JWMNP")
  }
  
  numeric_vars <- c(numeric_vars, "PWGTP")
  
  # check for valid categorical vars
  categorical_vars_opts <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  if(length(categorical_vars) == 0) stop("No categorical variable selected. Specify at least one or leave SEX as the default")
  for(i in categorical_vars) {
    if(!(i %in% categorical_vars_opts)) stop("Invalid categorical variable. Specify a single variable or a vector of variables. Options: FER, HHL, SCH, SCHL, SEX")
  }
  
  # check for valid geography scope and region of interest
  geography_opts <- c("Region", "State", "Division")
  if(!(geography %in% geography_opts)) stop("Invalid geography classification. Specify one of: State, Region, Division")
  if(specific_geo == "all") {
    specific_geo = "*"
  }
  else {
    if(geography == "State" & !(specific_geo %in% 1:50)) stop("Invalid state requested. Specify a number 1 to 50 or `all`")
    if(geography == "Region" & !(specific_geo %in% 1:4)) stop("Invalid region requested. Specify a number 1 to 4 or `all`")
    if(geography == "Division" & !(specific_geo %in% 1:9)) stop("Invalid division requested. Specify a number 1 to 9 or `all`")
  }
  
  api_key <- "59c3be738e0d051c63b57b6327de0b11ef0f7a1e"
  api_url <- paste0("https://api.census.gov/data/",
                    year,
                    "/acs/acs1/pums?get=",
                    paste0(numeric_vars, collapse = ","), # reformat numeric vars
                    ",",
                    paste0(categorical_vars, collapse = ","), # reformat categorical vars
                    "&for=",
                    tolower(geography), # region type
                    ":",
                    specific_geo, # specific region
                    "&key=",
                    api_key)
  
  api_return <- GET(api_url)
  api_parsed <- fromJSON(rawToChar(api_return$content)) |>
                as_tibble()
  
  # fix col names
  output <- api_parsed
  names(output) <- output[1,]
  output <- output[-1,]
  
  
  fctr_vars <- c("SEX", "FER", "HHL", "SCH", "SCHL")
  numeric_vars <- c("PWGTP", 'AGEP', "GASP", "GRPIP", "JWMNP")
  time_Vars <- c("JWAP", "JWDP")
  
  # recode variables
  output <- output |> 
            mutate(across(any_of(numeric_vars), ~ as.numeric(.x)),
                   across(any_of(fctr_vars), ~ as.factor(.x)))

  # return just a preview to test func
  return(output)
}
```

Double checking to see if our recoding works:
```{r}
head(query.census())
```













